A comparison of design-based and model-based approaches for finite population spatial data.

### 2021-10-21 inst/manuscript/manuscript.Rmd
minor discussion edits.

Add sentence about bias.
Add sentence about user functions.

@@

```{r, results = "hide"}
## GRTS Sample, Design Analysis
data_sf <- sf::st_as_sf(nla_nomiss_both, coords = c("xcoords", "ycoords"),
                        crs = 5070)
library(spsurvey)
grts_samp <- grts(data_sf, n_base = n)
grts_bind <- sp_rbind(grts_samp) # Matt you need to install newest version
# devtools::install_github("USEPA/spsurvey", ref = "main")

@@

The design-based and model-based approaches to inference are fundamentally different paradigms by which to select samples and analyze data. The design-based approach incorporates randomness through sampling to estimate a population parameter. The model-based approach incorporates randomness through distributional assumptions to predict the realized value of a random variable. Though these approaches have often been compared in the literature both from theoretical and analytical perspectives, our contribution lies in studying them in a spatial context while implementing spatially balanced sampling. Aside from the theoretical differences described, a few analytical findings were particularly notable: the design decision (GRTS vs IRS) seems much more important than the analysis decision (design-based vs model-based); Independent of the analysis approach, there is no reason to prefer IRS over GRTS for spatial data -- GRTS tends to perform at least as well as IRS when there is no spatial correlation increasingly than IRS as the strength of spatial correlation increases; the gap in relative efficiency between GRTS-design and GRTS-model widens as the strength of spatial correlation increases; and when the data are skewed, interval coverage for all approaches improves both as the sample size increases and as the strength of correlation increases.

There are several benefits and drawbacks of the design-based and model-based approaches for spatial data, some of which we have not yet discussed but are worthy of consideration in future research. The design-based approach relies on few assumptions, while the model-based approach relies on rigid distributional ones. The Horvitz-Thompson estimator of means and totals is unbiased and computationally efficient, while model-based estimation of covariance parameters can be computationally burdensome, especially for likelihood-based methods such as REML that rely on inverting a covariance matrix. The design-based approach also more naturally handles binary data, free from the more complicated logistic regression formulation commonly used to handle binary data in a model-based approach. The model-based approach, however, can quantify the relationship between covariates (predictor variables) and the response variable, something the design-based approach cannot do naturally. The model-based approach also yields estimated spatial covariance parameters, which help better understand the process of study. Model selection is also possible using model-based approaches and criteria such as likelihood ratio tests or AIC [@akaike1974new]. Model-based approaches are capable of more efficient small-area estimation than design-based approaches by leveraging distributional assumptions in areas with few observed sites. Model-based approaches can also compute site-by-site predictions at unobserved locations and use them to construct informative visualizations. The benefits and drawbacks of both approaches, alongside our theoretical and analytical comparisons, should be heavily considered when choosing among them. This is especially true from an analysis perspective, as we found that using a spatially balanced sampling algorithm benefits both design-based and model-based analyses.

>>>>

@@

```{r, results = "hide"}
## GRTS Sample, Design Analysis
data_sf <- sf::st_as_sf(nla_nomiss_both, coords = c("xcoords", "ycoords"),
                        crs = 5070)
library(spsurvey)
grts_samp <- grts(data_sf, n_base = n)
grts_bind <- sp_rbind(grts_samp)

@@

The design-based and model-based approaches to inference are fundamentally different paradigms by which to samples are selected and data are analyzed. The design-based approach incorporates randomness through sampling to estimate a population parameter. The model-based approach incorporates randomness through distributional assumptions to predict the realized value of a random variable. Though these approaches have often been compared in the literature both from theoretical and analytical perspectives, our contribution lies in studying them in a spatial context while implementing spatially balanced sampling. Aside from the theoretical differences described, a few analytical findings are particularly notable: the design decision (GRTS vs IRS) seems much more important than the analysis decision (design-based vs model-based); independent of the analysis approach, there is no reason to prefer IRS over GRTS for spatial data -- GRTS tends to perform at least as well as IRS when there is no spatial correlation and increasingly better than IRS as the strength of spatial correlation increases; the gap in relative efficiency between GRTS-design and GRTS-model widens as the strength of spatial correlation increases; and when the data are skewed, interval coverage for all approaches improves both as the sample size increases and as the strength of correlation increases.

<!-- The design-based approach relies on few assumptions, while the model-based approach relies on rigid distributional ones. -->

There are several benefits and drawbacks of the design-based and model-based approaches for spatial data, some of which we have not yet discussed but are worthy of consideration in future research.  Design-based approaches are often computationally efficient, while model-based estimation of covariance parameters can be computationally burdensome, especially for likelihood-based methods such as REML that rely on inverting a covariance matrix. The design-based approach also more naturally handles binary data, free from the more complicated logistic regression formulation commonly used to handle binary data in a model-based approach. The model-based approach, however, can more naturally quantify the relationship between covariates (predictor variables) and the response variable. The model-based approach also yields estimated spatial covariance parameters, which help better understand the process of study. Model selection is also possible using model-based approaches and criteria such as likelihood ratio tests or AIC [@akaike1974new]. Model-based approaches are capable of more efficient small-area estimation than design-based approaches by leveraging distributional assumptions in areas with few observed sites. Model-based approaches can also compute site-by-site predictions at unobserved locations and use them to construct informative visualizations. The benefits and drawbacks of both approaches, alongside our theoretical and analytical comparisons, should be heavily considered when choosing among them. This is especially true from an analysis perspective, as we found that using a spatially balanced sampling algorithm benefits both design-based and model-based analyses.
